---
title: "Course Project 1"
author: "Aditya Iyengar"
date: "09/04/2020"
output: html_document
---

This is the documentation for Course Project 1 for the online course **Reproducible Research** offered by *Johns Hopkins University* on Coursera.  
You can find the project details in the Readme file.  

### Step 1: Loading and Preprocessing the Data
Let's load the data from the link and unzip it.
```{r, echo = TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", "activity.zip", method = "curl")
dataset <- read.csv(unzip("activity.zip"))
```
Here's a basic overview of the dataset.
```{r, echo = TRUE}
str(dataset)
```
Let's also process the dates by converting them using the *lubridate* package.
```{r, echo = TRUE, results = "hide"}
library(lubridate)
dataset$date <- ymd(dataset$date)
```
  
### Step 2: Analysis of the number of steps taken
First, we calculate the total number of steps per day.
```{r, echo = TRUE}
dailysteps <- tapply(dataset$steps, dataset$date, sum, na.rm = TRUE)
```
We then calculate the mean and median steps taken per day. 
```{r, echo = TRUE}
print(paste("Mean steps per day:", (round(mean(dailysteps, na.rm = TRUE), digits = 2))))
print(paste("Median steps per day:", (round(median(dailysteps, na.rm = TRUE), digits = 2))))
```
Here is a histogram plot of the number of steps per day.
```{r, echo = TRUE}
hist(dailysteps, main = "Histogram of steps per day", xlab = "Total steps per day", ylab = "Number of days", col = "skyblue", ylim = c(0, 40))
```
  
### Step 3: Average Daily Activity Pattern
The time interval with the maximum steps per day, on an average is given by:
```{r, echo = TRUE}
stepstime <- tapply(dataset$steps, dataset$interval, mean, na.rm=TRUE)
print(paste(names(which(stepstime == max(stepstime))), "with", max(stepstime), "steps on an average"))
```
Here is a time series plot of the average daily activity trend.
```{r, echo = TRUE}
plot(names(stepstime), stepstime, type='l', main = "Daily Activity Trend", xlab = "Time", ylab = "Average steps", ylim = c(0, 250))
```
  
### Step 4: Imputing Missing Values
```{r, echo = TRUE}
print(paste("Number of missing values:", sum(is.na(dataset))))
```
To impute the missing values, we will use the mean steps during that particular time interval over the entire dataset.
```{r, echo = TRUE}
dataset1 <- dataset
dataset1 <- cbind(dataset, stepstime)
for(i in 1:length(dataset1$steps)){
  if(is.na(dataset1$steps[i])==TRUE){
    dataset1$steps[i] <- dataset1$stepstime[i] 
  }
}
```
Let us attempt to analyze this new dataset.
First, we calculate the total number of steps per day.
```{r, echo = TRUE}
dailysteps1 <- tapply(dataset1$steps, dataset1$date, sum, na.rm = TRUE)
```
We then calculate the mean and median steps taken per day. 
```{r, echo = TRUE}
print(paste("Mean steps per day:", (round(mean(dailysteps1, na.rm = TRUE), digits = 2))))
print(paste("Median steps per day:", (round(median(dailysteps1, na.rm = TRUE), digits = 2))))
```
We observe an increase in the mean and median steps per day. This occurs because the NA values are replaced by average imputed values.  
Here is a histogram plot of the number of steps per day.
```{r, echo = TRUE}
hist(dailysteps1, main = "Histogram of steps per day with imputed data", xlab = "Total steps per day", ylab = "Number of days", col = "skyblue", ylim = c(0, 40))
```
  
### Step 5: Weekends vs Weekdays
Let us compare the daily activity trend for weekends and weekdays.
```{r, echo = TRUE}
library(chron)
dataset2 <- dataset1
isweekend <- is.weekend(dataset2$date)
dataset2 <- cbind(dataset2, isweekend)
dataset3 <- dataset2
for(i in 17568:1){
  if(dataset2$isweekend[i]==TRUE){
    dataset2 <- dataset2[-i, ]
  }
}
for(i in 17568:1){
  if(dataset3$isweekend[i]==FALSE){
    dataset3 <- dataset3[-i, ]
  }
}
stepstime2 <- tapply(dataset2$steps, dataset2$interval, mean, na.rm=TRUE)
stepstime3 <- tapply(dataset3$steps, dataset3$interval, mean, na.rm=TRUE)
plot(names(stepstime2), stepstime2, main = "Weekday Activity", type = "l", xlab = "Time", ylab = "Average steps")
plot(names(stepstime3), stepstime3, main = "Weekend Activity", type = "l", xlab = "Time", ylab = "Average steps")
```